\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-22 pldi/draft.tex
Pass over implementation

%% \begin{definition}
%% Given a dataflow graph $G = (V, E, O)$, where $V$ is a set of nodes
%% representing commands, $E$ is a set of edges representing files, and
%% $O$ is a function from $V \cup v_{out}$ to a total order of incoming
%% edges. $v_{out}$ represents a concatenation of all the outputs of the
%% dataflow graph. We represent the total order for a node $v$ as
%% $<_v$. Given a node a node $v$
%% with input edges $ie = \{ i_1 = (v_{i_1}, v, 1), i_2, ..., i_n =
%% (v_{i_n}, v, n) \}$ and output edge $o = (v, v_o, a)$, we define a
%% complete node split $s(v, G) = (V', E')$ where $V' = V - v \cup \{
%% v_1, ..., v_n \}$ and $E' = E - ie \cup \{(v_{i_1}, v_1, 1), ...,
%% (v_{i_n}, v_n, 1) \} - o \cup \{ (v_1, v_o\}$
%% \end{definition}

@@

\section{From scripts to graphs and back again}
\label{impl}

\kk{Less fun section title: Implementation}

This section describes challenges related to the implementation of
\sys. It starts by describing how the \sys front-end translates a
shell script to the dataflow graph model defined in \S\ref{ir} by
identifying and transforming parallelizable regions~\sx{front-end}. It
closes with the \sys runtime\todo{/backend} and the technical
challenges that were addressed to achieve performance
benefits~\sx{optimizer}.

@@

As mentioned in \Cref{ir}, in order to optimize a shell script, \sys
translates it to a dataflow graph.  However, there are several
components in shell programs that cannot be arbitrarily parallelized
without affecting the program behaviour---\eg commands connected with
a \ttt{\&\&} operator~\sx{bg}.
%% In addition, the shell's highly dynamic
%% nature compounds the challenge---\eg by not knowing what files does an
%% unexpanded command argument refer to.
To address this issue, we introduce the notion of parallelizable
program regions and a translation pass for converting them to dataflow
graphs.

@@

\heading{Parallelizable Regions}
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Parallelizable regions are program sub-expressions that can be
parallelized safely without altering the total program behaviour.
%% respect to their sequential execution.
The search for these regions can be guided by the structure of shells
programs, which inherently contains information about components that
can be executed independently, and barriers that are natural points for
enforcing synchronization. Consider the following example:

@@

\noindent
The commands \ttt{cat f1 f2} and \ttt{grep "foo"} would execute as
independent processes in the standard shell, while \ttt{sort f3} would
wait for their completion before executing.  Both \ttt{cat f1 f2 |
  grep "foo" > f3} and \ttt{sort f3} are therefore parallelizable
regions, regions that cannot be extended beyond \ttt{\&\&}---even
better, they are maximal.

Intuitively, parallelizable regions correspond to sub-expressions of
the program that would be allowed to execute independently by
different processes in the POSIX standard~\cite{posix}. Larger
parallelizable regions can be composed from smaller ones using the
pipe operator \ttt{|} and the parallel composition operator \ttt{\&}.
Conversely, all other operators, such as the \ttt{;} sequential
composition operator and the \ttt{\&\&} logical operator, represent
barrier constructs that do not allow parallelizable regions to
permeate through.

@@

\heading{Translation Pass}
%
%% \kk{It might be a good idea to change this subsection based on the new
%%   as-late-as-possible architecture where everything is wrapped and
%%   translated at the final moment. Ideally, this would become the main
%%   point of this subsection, as the translation algorithm is kind of
%%   redundant when having the distributable regions too.}
%
The \sys front-end performs a depth-first search pass on the AST of
the given shell program.  During this pass, it extends the
parallelizable regions from the bottom up, translating their
independent components to dataflow graph nodes until a barrier
construct is reached. All subtrees that are not translated to dataflow
graphs are kept as they are. The output of the translation pass is the
original AST where parallelizable regions have been replaced with
dataflow graphs and a call to \sys's runtime.

To identify possible opportunities for parallelism, the translation
pass extracts the parallelizability class of each command together
with its inputs and outputs. It achieves that by searching all
available PAL~\sx{ext} records for each command and resorts to
conservative defaults if a record for the command is not found. If the
command is \sta or parallelizable \pur, the translation pass initiates
a parallelizable region that is propagated up the tree.

@@

\heading{Dangling FIFOs and Zombie Producers}
%
%% \TODO{(This problem is a nice followup for the one above) Problem:
%%   Dataflow graph termination. Illustration: Show the example of head
%%   deadlock and explain that this is inherent behaviour in the shell.
%%   (Solution: emptying pipelines or terminating processes) In order for
%%   things to terminate gracefully, dish wraps all commands in order to
%%   always consume all of their input by sending it down to
%%   /dev/null. Warning: This means that all commands in the pipeline
%%   must terminate gracefully if all of their output is consumed (if a
%%   process dies unexpectedly, noone will consume its input. We leave
%%   addressing that and other monitoring questions for future work). We
%%   evaluate the overhead of emptying all outputs in section X.}
%
% \nv{I think all of these titles should be the problem/challenge, not the solution}
% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
Under normal operation, a command exits after it has produced and
sent all of its output to
%% a pipe
its output channel.
% \ttt{SIGPIPE} 
However, if its output is a pipe and the pipe reader exits early, the
command needs to be notified so that it stops writing to it.
%% \todo{appending to the stream/trying to push data to its output
%%   channel}.
% ---so that it avoids unnecessary work and buffering challenges in the kernel.
In \unix, this is achieved by an out-of-band error mechanism: the
operating system will deliver a \ttt{PIPE} signal to the producer,
notifying it that the pipe's consumer has exited.  This is different
from errors for other system calls
%% return an error
and unusual compared to non-\unix systems\footnote{For example,
  Windows indicate errors for \ttt{WriteFile} using its return
  code---similar to \ttt{DeleteFile} and other Win32 functions.}
primarily because pipes and pipelines are at the heart of its
philosophy.
% write errors are communicated differently 
Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes\kk{I think that it is also possible for anonymous, e.g. \ttt{cat f1 - | head -n 1}}---\unix cannot signal this condition.
Consider the following emitted code from \sys's earlier versions:
%% \kk{Proposed alternative: ``Consider the following shell script:''}
\begin{lstlisting}[language=sh, numbers=none]
  mkfifo fifo1 fifo2
  cat file-chunk1.txt > fifo1 &
  cat file-chunk2.txt > fifo2 &
  cat fifo1 fifo2 | head -n 1 & wait
\end{lstlisting}
\noindent
In the code above, \ttt{head} exits early causing the last \ttt{cat}
to exit before opening \ttt{fifo2}.  As a result, the second \ttt{cat}
never receives a \ttt{PIPE} signal that its consumer exited---after
all, \ttt{fifo2} never even had a consumer.  This, in turn, leaves the
second \ttt{cat} unable to make progress, both (i) blocked unable to
write output, and (ii) unaware that its consumer has exited.  Coupled
with \ttt{wait} at the end, the entire snippet reaches a deadlock.

@@

\heading{Pure Fold Function Implementations}

As mentioned in \Cref{parallelizability}, commands in \tpur can be
parallelized using a map stage and a \todo{fold} stage. We have
implemented these stages for several commands in \tpur to enable \sys
to parallelize them. Some interesting examples are \ttt{sort}, which
can be parallelized simply by using \ttt{sort} as a map and \ttt{sort
  -m} as \todo{fold}, and \todo{wc} \TODO{Niko, can you complete
  that with wc or any other interesting example?}

>>>>

%% \begin{definition}
%% Given a DFG $G = (V, E, O)$, where $V$ is a set of nodes
%% representing commands, $E$ is a set of edges representing files, and
%% $O$ is a function from $V \cup v_{out}$ to a total order of incoming
%% edges. $v_{out}$ represents a concatenation of all the outputs of the
%% DFG. We represent the total order for a node $v$ as
%% $<_v$. Given a node a node $v$
%% with input edges $ie = \{ i_1 = (v_{i_1}, v, 1), i_2, ..., i_n =
%% (v_{i_n}, v, n) \}$ and output edge $o = (v, v_o, a)$, we define a
%% complete node split $s(v, G) = (V', E')$ where $V' = V - v \cup \{
%% v_1, ..., v_n \}$ and $E' = E - ie \cup \{(v_{i_1}, v_1, 1), ...,
%% (v_{i_n}, v_n, 1) \} - o \cup \{ (v_1, v_o\}$
%% \end{definition}

@@

\section{From Scripts to Graphs and Back Again}
\label{impl}

This section describes \sys's implementation.
It first details how \sys's front-end translates a shell script to the DFG model defined in \S\ref{ir} by identifying and transforming parallelizable regions~\sx{front-end}.
It then describes \sys's runtime\todo{/backend} and the technical challenges addressed to achieve performance benefits~\sx{optimizer}.

@@

To optimize a shell script, \sys translates it to a DFG~\sx{ir}.
As there are several program fragments that cannot be parallelized---\eg commands connected with \ttt{\&\&}~\S\ref{bg}---\sys introduce the notion of \emph{parallelizable program regions} and a translation pass for converting them to DFGs.
%% In addition, the shell's highly dynamic
%% nature compounds the challenge---\eg by not knowing what files does an
%% unexpanded command argument refer to.

@@

\heading{Parallelizable Regions}
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Parallelizable regions are program sub-expressions that can be parallelized safely, \ie without breaking the program.
%% respect to their sequential execution.
The search for these regions is guided by the shell language and the structure of a particular program.
These contain information about two types of components:
  fragments that can be executed independently and barriers that are natural synchronization points. % for enforcing synchronization.
Consider this script:

@@

\noindent
The \ttt{cat} and \ttt{grep} commands execute independently (and concurrently) in the standard shell, but \ttt{sort} waits for their completion prior to start.
Both \ttt{cat f1 f2 | grep "foo" > f3} and \ttt{sort f3} are thus parallelizable regions, each not extending beyond \ttt{\&\&}---\ie more precisely, they are maximal.

Intuitively, parallelizable regions correspond to sub-expressions of
the program that would be allowed to execute independently by
different processes in the POSIX standard~\cite{posix}. Larger
parallelizable regions can be composed from smaller ones using the
pipe operator (\ttt{|}) and the parallel-composition operator (\ttt{\&}).
Conversely, all other operators, such as the sequential composition operator (\ttt{;}) and the logical operators (\ttt{\&\&}, \ttt{||}), represent
barrier constructs that do not allow parallelizable regions to
permeate through.

@@

\heading{Translation Pass}
%
%% \kk{It might be a good idea to change this subsection based on the new
%%   as-late-as-possible architecture where everything is wrapped and
%%   translated at the final moment. Ideally, this would become the main
%%   point of this subsection, as the translation algorithm is kind of
%%   redundant when having the distributable regions too.}
%
The \sys front-end performs a depth-first search pass on the AST of
the given shell program.  During this pass, it extends the
parallelizable regions bottom-up, translating their
independent components to DFG nodes until a barrier
construct is reached. All subtrees not translatable to DFGs are kept as they are. The output of the translation pass is the
original AST where parallelizable regions have been replaced with
DFGs and a call to \sys's runtime.

To identify opportunities for parallelism, the translation
pass extracts the parallelizability class of each command together
with its inputs and outputs.  To achieve this for each command,
it searches all available annotation~\sx{ext} and resorts to
conservative defaults if no record is found. If the
command is \sta or parallelizable \pur, the translation pass initiates
a parallelizable region that is propagated up the tree.

@@

\heading{Dangling FIFOs and Zombie Producers}
%
%% \TODO{(This problem is a nice followup for the one above) Problem:
%%   Dataflow graph termination. Illustration: Show the example of head
%%   deadlock and explain that this is inherent behaviour in the shell.
%%   (Solution: emptying pipelines or terminating processes) In order for
%%   things to terminate gracefully, dish wraps all commands in order to
%%   always consume all of their input by sending it down to
%%   /dev/null. Warning: This means that all commands in the pipeline
%%   must terminate gracefully if all of their output is consumed (if a
%%   process dies unexpectedly, noone will consume its input. We leave
%%   addressing that and other monitoring questions for future work). We
%%   evaluate the overhead of emptying all outputs in section X.}
%
% \nv{I think all of these titles should be the problem/challenge, not the solution}
% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
Under normal operation, a command exits after it has produced and sent all of its output to its output channel.
However, if its output is a pipe and the pipe reader exits early, the command is notified to stop writing early.
In \unix, this is achieved by an out-of-band error mechanism: the
operating system will deliver a \ttt{PIPE} signal to the producer,
notifying it that the pipe's consumer has exited.  This is different
from errors for other system calls
%% return an error
and unusual compared to non-\unix systems\footnote{For example,
  Windows indicates errors for \ttt{WriteFile} using its return
  code---similar to \ttt{DeleteFile} and other Win32 functions.}
primarily because pipes and pipelines are at the heart of its
philosophy.
% write errors are communicated differently 
Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
% \kk{I think that it is also possible for anonymous, e.g. \ttt{cat f1 - | head -n 1}}
% \nv{TODO! Will think about this later}
Consider the following script:
% emitted code from \sys's earlier versions:
%% \kk{Proposed alternative: ``Consider the following shell script:''}
\begin{lstlisting}[language=sh, numbers=none]
  mkfifo fifo1 fifo2
  cat file-chunk1.txt > fifo1 &
  cat file-chunk2.txt > fifo2 &
  cat fifo1 fifo2 | head -n 1 & wait
\end{lstlisting}
\noindent
In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
This, in turn, leaves it unable to make progress, as it is is both blocked and unaware of its consumer exiting.
Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.

@@

\heading{Pure Fold Function Implementations}
Commands in \tpur can be parallelized using a \emph{map} and a \emph{fold} stage~\sx{parallelizability}.
\sys implements \emph{fold} for several commands in \tpur to enable parallelization.
A few interesting examples include \emph{fold} functions for
  (i) \ttt{sort}, which amounts to the merge phase of a merge-sort (and on GNU systems is implemented as \ttt{sort -m}),
  (ii) \ttt{uniq} and \ttt{uniq -c}, which need to check conditions at the boundary of their streams,
  (iii) \ttt{tac}, which consumes stream descriptors in the reverse order, and 
  (iv) \ttt{wc}, which adds inputs with an arbitrary number of elements (\eg \ttt{wc -lw} or \ttt{wc -lwc} \etc).
The combiners iterate over the provided stream descriptors, \ie they work with more than two inputs, and apply pure functions at the boundaries of input streams (with the exception of \ttt{sort} that has to interleave inputs).
% An important concern has been to maintain whitespace separation while
